---
output: github_document 
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

# ML2Pvae

This package allows for building, training, and evaluating ML2P-VAE parameter estimation models in Item Response Theory (IRT). These methods are based off of the work of Curi et al. (https://ieeexplore.ieee.org/document/8852333). Specific modifications are made to the VAE architecture which allow for interpretation of some of the trainable weights/biases, as well as the values of a hidden layer of the neural network.


## Installation

ML2Pvae can be installed using with the command
```{r}
#> library(devtools)
#> install_github("converseg/ML2Pvae")
```

## Example

We demonstrate the workflow of this package on a simulated dataset which consists of responses by 5000 subjects on an assessment with 30 items. It is assumed that the assessment evaluates 3 latent skills. This is a vanilla example, and much more detail and explanation is given to the examples found in the "vignettes" directory.

```{r}
library(ML2Pvae)

# Load sample data - included in package
data <- as.matrix(responses)
Q <- as.matrix(q_matrix)
```

First set a few hyper-parameters to build the neural network. Note that \textbf{num_items} and \textbf{num_skills} are fixed by the data set. The core function here is \textbf{build_vae_standard_normal()}, which constructs a modified neural network which can be used for parameter estimation.

```{r}
# Model parameters
num_items <- as.double(dim(Q)[2])
num_skills <- as.double(dim(Q)[1])
num_students <- dim(data)[1]
enc_arch <- c(16L, 8L)
enc_act <- c('relu', 'tanh')
out_act <- 'sigmoid'
kl <- 1

models <- build_vae_standard_normal(
  num_items,
  num_skills,
  Q,
  model_type = 2,
  enc_hid_arch = enc_arch,
  hid_enc_activation = enc_act,
  output_activation = out_act)
encoder <- models[[1]]
decoder <- models[[2]]
vae <- models[[3]]
```

Next, we set training parameters and fit the model to the data.

```{r, include=TRUE}
# Training parameters
num_train <- floor(0.8 * num_students)
num_test <- num_students - num_train
data_train <- data[1:num_train,]
data_test <- data[(num_train + 1):num_students,]
num_epochs <- 15
batch_size <- 8

# Train model
history <- train_model(
  vae,
  data_train,
  num_epochs = num_epochs,
  batch_size = batch_size,
  verbose = 1)
```

After the VAE has been trained, we can obtain IRT parameter estimates. Due to the modifications to the neural architecture, we are able to interpret the learned weights/biases in the VAE decoder as discrimination/difficulty parameter estimates. We can also feed forward student responses through the decoder to obtain estimates for student abilities.

```{r}
# Get IRT parameter estimates 
item_param_estimates <- get_item_parameter_estimates(
  decoder, model_type = 2)
diff_est <- item_param_estimates[[1]]
disc_est <- item_param_estimates[[2]]
test_theta_est <- get_ability_parameter_estimates(
  encoder, data_test)[[1]]
all_theta_est <- get_ability_parameter_estimates(
  encoder, data)[[1]]
```

Though real-world datasets likely won't have "true" values of the parameters available, our simulated dataset includes them for reference. Of course, none of these were used when we trained the model.

```{r}
# Load in true values (included in this package)
disc_true <- as.matrix(disc_true)
diff_true <- as.matrix(diff_true)
theta_true<- as.matrix(theta_true)
```
```{r, echo=FALSE}
# Examine model estimates
par(pty='s', mfrow=c(1,3))
plot(diff_true, diff_est, pch = '*',
       xlim = c(-3.1,3.1), ylim = c(-3.1,3.1),
       main = 'Difficulty Parameters',
       xlab = 'True', ylab = 'Estimated')
matplot(t(disc_true), t(disc_est), pch = '*',
       xlim = c(0.05, 2), ylim = c(0.05, 2),
       main = 'Discrimination Parameters',
       xlab = 'True', ylab = 'Estimated')
matplot(theta_true[4200:4400,], all_theta_est[4200:4400,], pch = '*',
       xlim = c(-3,3), ylim = c(-3,3),
       main = 'Ability Parameters',
       xlab = 'True', ylab = 'Estimated')
```

This was a very basic example, and results can be improved by fine-tuning the network architecture.

## References

\bibitem[curi] Curi et. al. “Interpretable Variational Autoencoders for Cognitive Models.” In Proceddings of the International Joint Conference on Neural Networks (IJCNN),2019.

\bibitem[converse] Converse, Curi, Oliveira. “Autoencoders for Educational Assessment.” In Proceedings of the Conference on Artifical Intelligence in Education (AIED), 2019.
